---
title: "Baby steps: running the test one line at a time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Baby steps: running the test one line at a time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

This vignette is for someone who has never used the package, and maybe has
not used R much either. There is no mathematics in it. Every step is shown,
and after each step you should look at what R printed and check that it
matches. If you want to know *why* the steps work, read the vignette called
"The mathematics behind the test" afterwards.

The question the package answers is this. A plant population reproduces
(flowers, fruits, seeds) and later new seedlings appear (recruits). If the
seedlings are simply the seeds from some earlier time, then a record of
reproduction and a record of seedlings should look like each other, shifted
in time. The package checks whether that is possible.

## Step 1. Load the packages

```{r, message = FALSE}
library(recruitlag)
library(dplyr)
library(ggplot2)
```

Nothing is printed. That is correct: loading a package is silent. `dplyr`
is for picking rows and columns out of tables, `ggplot2` for drawing.

## Step 2. Look at a data set

The package includes a made-up data set called `regimes`. It is one table
with six census records stacked in it, one row per record, unit (think
"host tree") and census period.

```{r}
regimes
```

The first line says it has 2,880 rows and 8 columns. We take one record
out of it, the one in which reproduction is steady through the year but the
seedlings arrive in bursts.

```{r}
r <- regimes |> filter(record == "aseasonal_gated")
r
```

`filter()` keeps only the rows of that record: 20 units times 24 periods is
480 rows. Each row has the reproduction counted on one unit at one period
and the recruits (new seedlings) seen there. Recruits are `NA` at periods 1
to 5, which means "not scored": the test only uses periods that have a full
lag window behind them.

## Step 3. Add up over units

The first test works on the whole population, so add the units together.

```{r}
totals <- r |>
  group_by(period) |>
  summarise(reproduction = sum(reproduction),
            recruits = if (all(is.na(recruits))) NA else sum(recruits))
totals
```

Look down the `reproduction` column: the numbers wander around a few hundred
and never do anything dramatic. Now look at `recruits`: mostly small numbers,
then three huge ones. That difference is what the test is about. Draw it:

```{r}
totals |>
  tidyr::pivot_longer(c(reproduction, recruits), names_to = "series", values_to = "count") |>
  ggplot(aes(period, count)) +
  geom_col(fill = "grey40") +
  facet_wrap(~ series, ncol = 1, scales = "free_y") +
  theme_minimal()
```

## Step 4. Measure how "bursty" each series is

The package measures burstiness with a number between 0 and 1 called the Gini
coefficient. 0 means every census had the same count. Close to 1 means almost
everything happened at one census.

```{r}
rain_gini(totals$reproduction)
rain_gini(totals$recruits)
```

(`rain_gini()` ignores the `NA`s.) Reproduction scores very low: it is spread
evenly. Recruits score high: they are concentrated in a few censuses. Write
those two numbers down.

## Step 5. Ask the key question

Could a delay turn the smooth series into the bursty one? The answer is
found with `lag_ceiling()`. Give it the table of totals and tell it how far
back in time a delay is allowed to reach (`K = 4` means five census
intervals). It finds the `period`, `reproduction` and `recruits` columns by
name.

```{r}
ce <- lag_ceiling(totals, K = 4, kernels = lag_kernels(4))
ce
```

Read the printout slowly. The line "reproduction (theorem bound)" repeats the
Gini of reproduction. That number is the *ceiling*: no delay, of any shape,
can make expected recruitment burstier than that. The line "observed
recruitment" is the Gini of recruits. The last line divides one by the
other: observed recruitment is many times burstier than any delay could make
it.

## Step 6. Draw it

```{r}
rain_strip(`aseasonal, gated` = ce)
```

The strip is a ruler from 0 to 1. The blue tick is the ceiling. The orange
dot is what was observed. The dot sits far to the right of the tick, in the
region no delay can reach.

## Step 7. See what a delay would actually predict

Here are four possible delays. The first one is what most population models
assume: seedlings appear one census after reproduction. The others spread the
delay in different ways.

```{r}
delays <- list(
  `one interval only` = c(1, 0, 0, 0, 0),
  `early`             = c(.7, .2, .07, .03, 0),
  `evenly spread`     = rep(.2, 5),
  `late`              = c(0, .03, .07, .2, .7))
rain_profiles(delays)
```

Now apply each of them to the real reproduction series and compare with the
recruits that were counted:

```{r}
rain_expected(ce, delays)
```

The bars are the seedlings counted. The lines are what each delay predicts.
All four lines are nearly flat and nearly on top of one another, and none of
them goes anywhere near the three tall bars. This is the picture that the
numbers in Step 5 summarise.

## Step 8. The careful version of the test

Step 5 compared the *average* that a delay predicts with the *counts* that
were observed. Counts are noisy, and a noisy count series can look bursty
even when its average is flat. The careful test takes that into account and
also uses the fact that recruits were counted separately on 20 units. It
works from the unit-by-period table `r` directly. It takes a few seconds
because it simulates many count series.

```{r}
set.seed(1)
ht <- host_lag_test(r, K = 4, nsim = 2000, kernels = lag_kernels(4))
ht
```

The last lines are the verdict. For each way of measuring burstiness the test
tried every delay it knows and reports the *best case for the delay
hypothesis*: the largest probability, over all delays, of getting a record as
bursty as the one observed. If even the best case is a very small number, no
delay can explain the record. The full table, one row per delay, is in
`ht$table`:

```{r}
ht$table |> select(label, phi, p_silent, p_gini, p_cv) |> arrange(desc(p_gini)) |> head()
```

## Step 9. Check yourself with a record where a delay is the truth

The same data set has a record in which recruits really are a delayed
consequence of reproduction. Repeat Steps 2, 3, 5 and 8 on it.

```{r}
r2 <- regimes |> filter(record == "aseasonal_delayed")
ce2 <- lag_ceiling(r2, K = 4, kernels = lag_kernels(4))   # a unit-by-period table is summed for you
ce2
set.seed(1)
ht2 <- host_lag_test(r2, K = 4, nsim = 2000, kernels = lag_kernels(4))
ht2
```

Two things to notice. First, in the ceiling printout the observed Gini is
again above the ceiling: counts are noisier than averages, so this on its own
proves nothing (that is exactly why Step 8 exists). Second, in the careful
test the best-case probabilities are now ordinary numbers, not tiny ones: the
careful test does not reject a delay when a delay is the truth.

## Step 10. The other way round: a seasonal plant

Everything so far used a plant that reproduces steadily all year. Many
plants do not: they flower in one season. Repeat Steps 2 to 5 on such a
record and watch what changes.

```{r}
r3 <- regimes |> filter(record == "seasonal_delayed")
totals3 <- r3 |> group_by(period) |>
  summarise(reproduction = sum(reproduction),
            recruits = if (all(is.na(recruits))) NA else sum(recruits))
totals3$reproduction
```

Large numbers for a few periods, then almost nothing, then large again a
year later. That is seasonal reproduction. Measure it:

```{r}
rain_gini(totals3$reproduction)
rain_gini(totals3$recruits)
```

Reproduction now scores high on the burstiness scale, because it is
concentrated in a season, and recruits score a little *lower* than
reproduction. Now the ceiling:

```{r}
ce3 <- lag_ceiling(totals3, K = 4, kernels = lag_kernels(4))
ce3
rain_strip(`aseasonal, gated` = ce, `seasonal, true delay` = ce3)
```

In the strip for the seasonal plant the orange dot sits to the *left* of the
blue tick, inside the region a delay can reach. So a delay could have
produced this record, and in this case one did: this record was built with a
true delay.

Here is the important part, so read it twice. A dot to the left of the tick
does not prove there was a delay. It only says the test cannot rule one out.
Try the seasonal record with the gate instead:

```{r}
r4 <- regimes |> filter(record == "seasonal_gated")
ce4 <- lag_ceiling(r4, K = 4, kernels = lag_kernels(4))
ce4
rain_strip(`seasonal, true delay` = ce3, `seasonal, gated` = ce4)
```

The gated record lands almost on the tick. With a seasonal plant, the true
delay and the gate look nearly the same on the ruler, because the ceiling is
so high that almost anything fits under it. The careful test of Step 8 does
not help much here either:

```{r}
set.seed(1)
host_lag_test(r4, K = 4, nsim = 2000, kernels = lag_kernels(4))
```

The best-case probabilities are ordinary numbers, even though we know this
record was gated. That is not a failure of the test; it is the honest
answer. When reproduction is seasonal, a burst of seedlings can always be
explained as "the seeds from last season", and no census record can say
otherwise.

## Step 11. What the four cases teach

You have now seen all four combinations. Keep this table in mind when you
read your own results.

| Reproduction | Recruits | Dot on the ruler | Careful test (Step 8) |
|---|---|---|---|
| steady all year | bursts (gated) | far right of the tick | tiny probabilities: no delay can explain it |
| steady all year | true delay | right of the tick, but only because counts are noisy | ordinary probabilities: a delay is possible |
| seasonal | true delay | left of the tick | ordinary: a delay is possible |
| seasonal | bursts (gated) | near the tick | ordinary: cannot tell |

The test can only say "no delay can explain this" when reproduction is
steady. When reproduction is seasonal, it cannot say much either way. That
is the whole reason the *Lepanthes* orchid in the `lepanthes_*` data sets is
a good case: it flowers every month of the year.

## Step 12. Your own data

Put your census record in a table with one row per unit (host tree, plot,
population) and period, and these columns: `unit`, `period` (an integer
starting at 1), `reproduction` and `recruits` (with `NA` at periods where
recruits were not scored). If your columns have other names, tell the
functions with `unit = `, `period = `, `reproduction = ` and `recruits = `,
as the *Lepanthes* vignette does. Then repeat Steps 3 to 8 with your table
in place of `r`, and read your result against the table in Step 11. If your
first lagged bin should be the census itself rather than the previous
census, add `lag0 = 0` to `lag_ceiling()` and `host_lag_test()`.
