---
title: "Reading the intervals: confidence, credible, and the four ways to cut one"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading the intervals: confidence, credible, and the four ways to cut one}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 3.6)
```

```{r, message = FALSE}
library(recruitlag)
```

The package reports the exceedance, observed concentration over the
ceiling, with an interval around it, and it offers three ways of making
that interval. They give numbers that are usually close and sentences that
are not interchangeable. This vignette says, as plainly as it can, what each
interval is, what it is not, where each is strong and weak, and how to write
the sentence that goes with it. It assumes no statistics beyond knowing what
a percentile is.

## The one distinction that matters

Every interval here answers one of two questions, and they are different
questions.

**A confidence interval** answers: *if the same study were repeated many
times, drawing a new set of units each time, and the interval were computed
each time by this recipe, how often would the interval contain the true
value?* The answer is the level, 90% here. The interval you have is one of
those repetitions. It either contains the true value or it does not, and
the 90% is a property of the recipe, not of this interval. That is the whole
of it, and it is why the sentence "there is a 90% probability that the
exceedance lies between 2.1 and 3.2" is not licensed by a confidence
interval, however natural it sounds.

**A credible interval** answers: *given the data and a stated prior, what is
the probability that the true value lies in this range?* The answer is
again the level, and now it is a probability about the value itself. The
sentence above is licensed. The price is the prior: a credible interval is
only as meaningful as the prior it was computed under, and the prior has to
be named.

The package labels every interval it returns, in the printed summary and in
the `interval` and `method` columns of `ceiling_draws_table()`, so that the
reader never has to guess which question was answered.

## The three routes

```{r, message = FALSE, warning = FALSE}
set.seed(20260913)
k <- lag_kernels(4, bin = 6, unit = "mo")
b  <- lag_ceiling_boot(lepanthes_hosts, unit = "host", reproduction = "inflorescences",
                       K = 4, kernels = k, R = 2000)
bb <- lag_ceiling_bayesboot(lepanthes_hosts, unit = "host", reproduction = "inflorescences",
                            K = 4, kernels = k, draws = 4000)
knitr::kable(ceiling_draws_table(`cluster bootstrap` = b, `Bayesian bootstrap` = bb)[
  , c("record", "quantity", "estimate", "lower", "upper", "interval", "probability", "probability_is")],
  digits = 2)
```

### The cluster bootstrap: a confidence interval

`lag_ceiling_boot()` takes the units of the study (host trees here), draws a
sample of the same size from them **with replacement**, so that some units
appear twice and some not at all, rebuilds both the reproductive series and
the recruit series from that sample, and recomputes the exceedance. It does
this 2,000 times. The spread of those 2,000 values is the bootstrap
distribution, and the interval is cut from it.

*What it assumes.* That the units are a sample from a larger population of
units like them, and that they are independent of one another. It assumes
nothing about the shape of the counts.

*Where it is strong.* No model, no prior, no distributional assumption. The
recipe is transparent: anyone can see exactly what was resampled.

*Where it is weak.* It needs enough units. With five sites the bootstrap
distribution has only a handful of distinct values, the interval is coarse,
and the bias-corrected version can fail; the function warns below ten units
and flags a failed row. It also cannot be applied to a series that has
already been pooled over units, because there is nothing left to resample.

*The sentence.* "The exceedance was 2.56 (90% confidence interval 2.1 to
3.2, cluster bootstrap over host trees)."

### The Bayesian bootstrap: a credible interval, same quantity

`lag_ceiling_bayesboot()` does not resample the units. It gives each unit a
random weight, drawn so that the weights are positive and sum to one (a
Dirichlet(1, ..., 1) draw), rebuilds both series as weighted sums, and
recomputes the exceedance; 4,000 times. This is Rubin's (1981) Bayesian
bootstrap, and it is a posterior in the strict sense: the prior is a flat
Dirichlet on the unknown distribution of units, which is the least
informative prior available for that object, and the weights are draws from
the posterior under it.

*What it assumes.* The same as the cluster bootstrap, plus that flat prior.
Because the prior is the same for every data set and puts no structure on
the counts, it is rarely the thing a reader objects to.

*Where it is strong.* The sentence may say "probability". `P(exceedance >
1)` is a posterior probability, and the interval is a credible interval.
Numerically it tracks the cluster bootstrap closely, because it is a
smoothed version of the same idea, so nothing is lost in exchange.

*Where it is weak.* Same as the cluster bootstrap on the number of units,
though it degrades more gracefully because the weights are continuous. And
the reader who does not accept Bayesian probability statements will read the
interval as if it were a confidence interval, which is approximately right
here and not something the package can prevent.

*The sentence.* "The exceedance was 2.56 (90% credible interval 2.1 to 3.0,
Bayesian bootstrap over host trees); the posterior probability that it
exceeds 1 is above 0.999."

### The negative-binomial model: a credible interval, different quantity

`lag_ceiling_stan()` fits a model to the recruit counts on every unit at
every period: each count is negative binomial, with a unit effect, a period
effect and a clumping parameter. From the posterior it takes the *expected*
recruit series, the series the counts scatter about, and puts that through
the ceiling arithmetic. The reproductive record enters as observed.

*What it assumes.* The model: negative-binomial counts, effects additive on
the log scale, and priors on the effect scales and the clumping. The priors
are those of the companion paper and are named in the help page; they are
working priors, and both are arguments.

*Where it is strong.* It estimates the object the ceiling actually bounds.
A count series is always more concentrated than its own expectation, so the
two bootstraps put an interval on a quantity that is inflated by sampling
noise. This route removes that inflation and asks whether what remains
still exceeds the bound. It is therefore the conservative reading, and the
gap between its estimate and the bootstraps' is itself a number worth
reporting: it is how much of the raw exceedance the counts alone could have
produced.

*Where it is weak.* It is a model, and it has to be checked: the function
prints the sampler diagnostics (divergent transitions, Rhat, effective
sample size) and warns when they are poor, and an interval from a poorly
sampled posterior is not an interval. It needs `cmdstanr` and CmdStan
installed. It refuses a reproductive rate, since the recruits must be
counts. And because it estimates a different quantity, its interval is not
a check on the other two: agreement is not expected, and disagreement is not
a fault.

*The sentence.* "The concentration of the expected recruit series exceeded
the ceiling by a factor of 2.1 (90% credible interval 1.6 to 2.8,
negative-binomial model with host and census effects); the posterior
probability that it exceeds 1 is 0.99." (Illustrative numbers.)

## The four ways to cut an interval from a set of draws

Once a route has produced its 2,000 or 4,000 values, an interval is cut
from them, and there are four recipes. Two are for the bootstrap and two are
for posteriors; the package offers both of each and defaults to one.

**Percentile.** The 5th and 95th percentiles of the values. The simplest
recipe, and exact when the bootstrap distribution is centred on the estimate
and symmetric. It is invariant to a change of scale: the 5th percentile of
the log exceedance is the log of the 5th percentile of the exceedance, so it
does not matter which scale you cut on.

**BCa** (bias-corrected and accelerated; Efron 1987). The percentile
interval with two corrections: one for the bootstrap distribution not being
centred on the estimate (bias), one for its skew (acceleration, estimated
from the leave-one-unit-out values). A ratio such as the exceedance is
skewed, so BCa is the default for the cluster bootstrap. It is also
invariant to a change of scale. Its cost is that with very few units the
corrections have too little to work with and the interval can collapse or
miss the estimate; the package flags that row instead of printing it.

**Equal-tailed** (ETI). The percentile recipe applied to posterior draws: the
central 90% of the posterior, 5% in each tail. Invariant to a change of
scale, for the same reason as the percentile interval. The default for the
two Bayesian routes.

**Highest-density** (HDI). The shortest interval that holds 90% of the
posterior. On a skewed posterior it is shorter than the equal-tailed
interval and sits closer to the mode. Its drawback is the one that matters
for a ratio: it is **not** invariant to a change of scale. The shortest 90%
of the exceedance and the shortest 90% of its logarithm, transformed back,
are different intervals, and the difference is visible on these data:

```{r}
ex <- bb$draws[, "exceedance_gini"]
eti <- quantile(ex, c(0.05, 0.95))
hdi <- function(v, level = 0.9) { v <- sort(v); n <- length(v); k <- floor(level * n)
  w <- v[(k + 1):n] - v[1:(n - k)]; i <- which.min(w); c(v[i], v[i + k]) }
rbind(`equal-tailed, on the ratio`       = eti,
      `equal-tailed, on log, back`       = exp(quantile(log(ex), c(0.05, 0.95))),
      `highest-density, on the ratio`    = hdi(ex),
      `highest-density, on log, back`    = exp(hdi(log(ex))))
```

The two equal-tailed rows agree to the last digit, as they must. The two
highest-density rows do not. An HDI is therefore a choice of scale as well
as of recipe, and the package offers it (`interval = "hdi"`) so that a user
who wants it can have it, with this paragraph as the reason it is not the
default.

## The two probability statements, side by side

The cluster bootstrap reports `p_boot`, the fraction of replicates in which
the exceedance was at most 1. The Bayesian routes report `prob`, the
fraction of posterior draws in which it exceeded 1. In a table the two
numbers can look like complements of each other. They are not the same kind
of number.

`p_boot` is a **one-sided bootstrap p-value**: the achieved significance
level of a test of "the record is no more concentrated than a delay
predicts". A small value is evidence against that hypothesis. It is not the
probability that the hypothesis is true, and the sentence "the probability
that the exceedance is at most 1 is 0.0005" is wrong. The printed summary
says so under the number, because the misreading is easy and common. It
also cannot be resolved below one over the number of replicates, so 0 is
printed as "< 1/R".

`prob` is a **posterior probability**, and the sentence "the probability
that the exceedance exceeds 1 is 0.999, under a flat prior on the
distribution of hosts" is exactly what it means.

If you want the second sentence, use the route that is built to give it.
Do not take the bootstrap's `p_boot`, subtract it from one, and call the
result a probability.

## Choosing, and reporting

| you want | use | interval | says |
|---|---|---|---|
| no model, no prior, the standard recipe | `lag_ceiling_boot()` | BCa (or percentile), a confidence interval | "90% CI" |
| a probability statement about the same quantity, minimal assumptions | `lag_ceiling_bayesboot()` | equal-tailed (or HDI), a credible interval | "90% CrI", `P(> 1)` |
| the quantity the bound applies to, with sampling noise removed | `lag_ceiling_stan()` | equal-tailed (or HDI), a credible interval | "90% CrI", `P(> 1)`, and the sampler diagnostics |

Whichever you choose, report the route and the recipe with the numbers,
because the reader cannot tell them apart from the digits: "2.56 (90% BCa
CI 2.1 to 3.2, cluster bootstrap over 23 host trees)" carries everything
needed to read it, and "2.56 (2.1 to 3.2)" does not. `ceiling_draws_table()`
carries those labels in its columns, and `rain_forest()` draws the routes
side by side so that a disagreement between them, which is where the
reading gets interesting, is seen rather than searched for.

```{r, fig.height = 2.6}
rain_forest(`L. eltoroensis, 23 hosts` = b, `L. eltoroensis, 23 hosts` = bb, index = "both")
```

## References

Efron, B. (1987) Better bootstrap confidence intervals. *Journal of the
American Statistical Association* 82: 171-185.

DiCiccio, T. J. and Efron, B. (1996) Bootstrap confidence intervals.
*Statistical Science* 11: 189-228.

Davison, A. C. and Hinkley, D. V. (1997) *Bootstrap Methods and their
Application*. Cambridge University Press.

Rubin, D. B. (1981) The Bayesian bootstrap. *The Annals of Statistics* 9:
130-134.

Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A. and
Rubin, D. B. (2013) *Bayesian Data Analysis*, 3rd edn. Chapman and Hall.
