---
title: "Invented records: where the test has power and where it does not"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Invented records: where the test has power and where it does not}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

```{r, message = FALSE}
library(recruitlag)
library(dplyr)
library(tidyr)
library(ggplot2)
```

The `regimes` tibble holds six constructed census records, one row per
record, unit and period: three reproductive regimes (seasonal, intermediate,
aseasonal) crossed with two recruitment mechanisms. Under `delayed`,
recruits on each of 20 units are a true delayed consequence of that unit's
own reproduction, through a fixed five-bin lag profile with
negative-binomial counts. Under `gated`, recruits arrive only at three of 24
periods, when an external gate opens, at a rate proportional to each unit's
average reproduction, so the record is episodic and does not track
reproduction. Nothing here is data; the generating rules are in
`data-raw/regimes.R` and in `regimes_truth`.

```{r}
regimes
regimes_truth
```

The point of the set is to show three things: the ceiling is high where
reproduction is seasonal and low where it is aseasonal; the pooled
comparison of observed concentration with the ceiling cannot decide the
question on its own; and the host-level test separates a true delay from a
gate where the ceiling is low, and loses that power as reproduction becomes
seasonal.

## The six records

System totals by period, then reproduction and recruits side by side, each
on its own scale from zero (recruits are an order of magnitude fewer and
would vanish on a shared axis).

```{r}
totals <- regimes |>
  group_by(record, label, period) |>
  summarise(reproduction = sum(reproduction),
            recruits = if (all(is.na(recruits))) NA_real_ else sum(recruits),
            .groups = "drop")
totals
```

```{r, fig.height = 8}
totals |>
  pivot_longer(c(reproduction, recruits), names_to = "series", values_to = "count") |>
  mutate(series = factor(series, c("reproduction", "recruits")),
         label = factor(label, unique(regimes$label))) |>
  mutate(panel = factor(paste(label, series, sep = ": "),
                        paste(rep(unique(regimes$label), each = 2), c("reproduction", "recruits"), sep = ": "))) |>
  ggplot(aes(period, count, fill = series)) +
  geom_col(width = 0.7, na.rm = TRUE) +
  facet_wrap(~ panel, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = c(reproduction = "#0072B2", recruits = "#D55E00"), guide = "none") +
  theme_minimal(base_size = 8.5) +
  theme(strip.text = element_text(hjust = 0))
```

Going down the left column, reproduction goes from strongly seasonal to
flat. In the right column, the `delayed` records track reproduction as a
smoothed, shifted copy, and the `gated` records arrive in bursts at the same
three periods whatever the reproductive regime.

## The ceiling

For each record, the ceiling is the concentration of reproduction itself,
and `lag_ceiling()` confirms by search that no lag profile exceeds it on the
scored periods. A unit-by-period table is summed to the system series
automatically, so each record can be passed straight in.

```{r}
k4 <- lag_kernels(4)
ceilings <- regimes |>
  group_split(record) |>
  setNames(sort(unique(regimes$record))) |>
  lapply(lag_ceiling, K = 4, kernels = k4)

tibble(record = names(ceilings),
       ceiling = sapply(ceilings, function(x) x$theorem["gini"]),
       observed = sapply(ceilings, function(x) x$obs["gini"]),
       exceedance = sapply(ceilings, function(x) x$exceedance["gini"])) |>
  mutate(across(where(is.numeric), round, 2))
```

```{r, fig.height = 5}
rain_strip(ceilings)
```

Two rows of the table carry the lesson. Under seasonal reproduction the
ceiling is about 0.65, and both records, the true delay and the gate, sit
near it: the gated record exceeds the ceiling only slightly, and a reader
could not tell the two mechanisms apart from the pooled series. Under
aseasonal reproduction the ceiling is near zero, so the gated record exceeds
it by an enormous factor, but so does the true delay, by a large one. That
second fact is the reason the pooled comparison cannot be the test: counts
are noisier than their expectation, and when the expectation is nearly
constant, any count series at all looks concentrated by comparison.

## Why the pooled comparison has no power

`ceiling_calibration()` simulates negative-binomial counts about two means
that both obey the ceiling, a flat mean and the best-kernel mean, and asks
how often the simulated series is at least as concentrated as the observed
one.

```{r}
set.seed(1)
cal <- lapply(ceilings, ceiling_calibration, nsim = 3000)
bind_rows(cal, .id = "record") |> select(record, null, phi, p_gini, p_cv)
```

Every probability is ordinary. The calibration says that a mean obeying the
ceiling, once sampled as clumped counts, routinely produces series as
concentrated as any of the six, gated or not. Pooled over units, the test
cannot refute anything.

## The host-level test

The null stated at the level at which the lag hypothesis is asserted:
expected recruitment on each unit is that unit's own lagged reproduction,
scaled to its own total, with clumping estimated by the moment identity
under that heterogeneous mean. Simulated counts are summed over units and
compared with the observed census totals on four statistics.

```{r}
set.seed(2)
tests <- regimes |>
  group_split(record) |>
  setNames(sort(unique(regimes$record))) |>
  lapply(host_lag_test, K = 4, nsim = 2000, kernels = k4)

verdict <- bind_rows(lapply(tests, `[[`, "verdict"), .id = "record") |>
  select(record, statistic, max_p) |>
  pivot_wider(names_from = statistic, values_from = max_p, names_prefix = "P_")
verdict
```

Read the rows in pairs. Under seasonal reproduction the true delay is not
refuted (P near 1), but neither is the gate: a lagged seasonal input itself
predicts a concentrated census series, and counts about that expectation are
as concentrated as the gated record, so the host-level test has little power
where the ceiling is high, exactly as the pooled comparison had none. Under
intermediate seasonality the true delay is not refuted and the gate is
refuted on the Gini and CV of the census totals, because its bursts do not
fall where the lagged peaks predict them. Under aseasonal reproduction the
true delay is again not refuted, with P in the middle of its range, and the
gate is refuted decisively. The silent-census statistic is uninformative in
these records because the gate keeps a small background rate open at every
period; in the *Lepanthes* record it is the most interpretable statistic
because the gate there closed completely.

```{r}
plot(tests$aseasonal_gated, statistic = "gini")
```

## Is the host-level test calibrated?

When the true profile is among the kernels, the probability it receives
should be uniformly distributed over replicate draws of the data. The
following draws new data sets from the aseasonal generating rule, evaluates
the true profile on each with a small simulation budget, and shows the
distribution of P. It is kept small so that the vignette builds quickly;
increase `n_rep` and `nsim` to tighten it.

```{r}
set.seed(11)
H <- 20; Tn <- 24; K <- 4; t_R <- 6:24
w_true <- regimes_truth |> filter(mechanism == "delayed") |> pull(value)
base <- round(exp(rnorm(H, log(12), 0.6)))
n_rep <- 12
p_true <- replicate(n_rep, {
  d <- tibble(unit = rep(sprintf("u%02d", 1:H), each = Tn), period = rep(1:Tn, H),
              reproduction = as.vector(t(sapply(base, function(b) rpois(Tn, b)))))
  d <- d |> group_by(unit) |>
    mutate(mu = c(rep(NA, 5), 0.08 * convolve_lag(reproduction, w_true, t_R, 1)),
           recruits = ifelse(is.na(mu), NA, rnbinom(n(), size = 2, mu = ifelse(is.na(mu), 0, mu)))) |>
    ungroup()
  ht <- host_lag_test(d, K = K, nsim = 300, flat = FALSE,
                      kernels = lag_kernels(K, families = "delay", extra = list(truth = w_true)))
  ht$table$p_gini[ht$table$label == "truth"]
})
round(sort(p_true), 2)
```

The values spread across the unit interval rather than piling up near zero,
which is what a calibrated test does when the null is true. A single draw
can of course land in the tail; the seed of `data-raw/regimes.R` was chosen
so that the delayed records in `regimes` are typical draws rather than tail
ones.

## What to take from the set

The ceiling is a statement about expectations, and where reproduction is
seasonal it is too high to bite. Where reproduction is aseasonal the ceiling
is low, but the pooled comparison still cannot decide the question, because
counts about a flat mean look concentrated. The decision comes from
replicate units: a true delay predicts, for each unit, a series shaped by
that unit's own reproduction, and a gate does not.
