---
title: "The Lepanthes eltoroensis analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Lepanthes eltoroensis analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

```{r, message = FALSE}
library(recruitlag)
library(dplyr)
```

This vignette reproduces the analysis of the companion paper on the
epiphytic orchid *Lepanthes eltoroensis* (Luquillo Mountains, Puerto Rico),
in which reproduction is continuous in time and recruitment is episodic. The
package's tests check the numbers below against the archived outputs of the
paper's pipeline; here they are simply computed.

## The two records

```{r}
lepanthes_monthly
lepanthes_census |> select(period, census, adults, inflorescences, recruits)
lepanthes_hosts
```

The monthly survey (1994 to 1996) counts inflorescence-bearing adults and new
seedlings at 23 monthly censuses of four populations; recruits are defined by
first appearance, so they are scored from census 2. The six-monthly survey
(1999 to 2004) counts every plant on 23 host trees at 12 censuses; recruits
are scored at censuses 3 to 12. `lepanthes_census` holds its system totals
and `lepanthes_hosts` one row per host tree and census, which is the input
of the host-level test.

The lag conventions differ, and the package makes them explicit. The
monthly lag model of the paper reads lag *k* at month *t - k* over a
six-month horizon (`K = 6`, `lag0 = 0`). The six-monthly model reads lag
*k* at census *t - 1 - k* over five bins spanning 6 to 30 months
(`K = 4`, `lag0 = 1`).

## Concentration and the ceiling

```{r}
k_m <- lag_kernels(6, bin = 1, unit = "mo")
k_6 <- lag_kernels(4, bin = 6, unit = "mo")
c_m <- lag_ceiling(lepanthes_monthly, reproduction = "reproductive_adults", K = 6, lag0 = 0, kernels = k_m)
c_6 <- lag_ceiling(lepanthes_census, reproduction = "inflorescences", K = 4, lag0 = 1, kernels = k_6)
c_m
c_6
```

At both time scales the ceiling is low because the input is flat, the
searched maximum is attained by a pure delay and lies within 0.01 of the
theorem bound (the small difference arises from backfilling pre-record
lags), and observed recruitment exceeds the ceiling by a factor of about 10
(monthly) and 2.5 (six-monthly).

```{r}
rain_strip(`monthly 1994` = c_m, `six-monthly 1999-2004` = c_6)
```

## What any lag profile predicts

```{r}
profiles <- list(
  `one interval only (projection-matrix null)` = c(1, 0, 0, 0, 0),
  `concentrated early (6 to 12 months)` = c(.7, .2, .07, .03, 0),
  `evenly spread (6 to 30 months)` = rep(.2, 5),
  `concentrated late (24 to 30 months)` = c(0, .03, .07, .2, .7),
  `fitted posterior mean` = lepanthes_profile$weight)
rain_profiles(profiles, bin = 6, unit = "months")
rain_expected(c_6, profiles)
```

The expected series all but coincide and follow the slow decline of
reproduction; none resembles the recruits counted (62 at census 5, none at
censuses 7, 8 and 12, 32 at census 10). The fitted posterior-mean profile
from the paper's distributed-lag model is among them and is
indistinguishable from the rest, which is the identification problem the
paper describes: with a flat input the lag cannot be recovered, however long
the series.

## Why the pooled comparison cannot decide

```{r}
set.seed(20260903)
ceiling_calibration(c_6)
```

About a mean that obeys the ceiling, negative-binomial counts with the
observed variance-to-mean ratio produce a series as concentrated as the
observed one roughly one time in five. The exceedance of the ceiling by the
pooled counts is therefore not evidence; the test has to be made across
host trees.

## The host-level test

```{r}
lepanthes_hosts |>
  filter(!is.na(recruits)) |>
  group_by(period) |>
  summarise(recruits = sum(recruits), hosts_recruiting = sum(recruits > 0))
```

Three of the ten censuses are silent (no recruit on any of the 23 hosts) and
one carries 62. Under the lag hypothesis, expected recruitment on each host
is that host's own lagged inflorescence record, scaled to its own total.
The kernels are the pure delays, every uniform window, geometric decay and
the fitted profile; the paper adds 800 Dirichlet draws, which take longer
and change nothing.

```{r}
set.seed(20260910)
ht <- host_lag_test(lepanthes_hosts, unit = "host", reproduction = "inflorescences",
                    K = 4, nsim = 4000,
                    kernels = lag_kernels(4, bin = 6, unit = "mo",
                                          extra = list(`fitted posterior mean` = lepanthes_profile$weight)),
                    phi = 0.089)
ht
```

The moment estimate of the clumping parameter is about 0.19; the paper also
evaluates every profile at 0.089, the estimate obtained about a single
pooled mean, which attributes the between-host spread of the means to
clumping and is about twice the clumping the counts support. Even at that
value the most favourable profile does not make three silent censuses
ordinary.

```{r}
plot(ht, statistic = "silent")
plot(ht, statistic = "gini")
```

## The verdict, keyed on labels

```{r}
ht$verdict
ht$table |>
  filter(phi_source == "moment",
         label %in% c("flat reference", "pure delay 24-30 mo", "uniform 0-30 mo", "fitted posterior mean")) |>
  select(label, phi, gini_expected, p_silent, p_gini, p_cv)
```

No lag profile, of any shape over 6 to 30 months, makes the observed record
ordinary. Recruitment in this population is not the delayed consequence of
its own reproduction; something else opens and closes.
