---
title: "The mathematics behind the test, and the functions that implement it"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The mathematics behind the test, and the functions that implement it}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r}
library(recruitlag)
```

## 1. The hypothesis as a convolution

Let $X_t$ be the reproductive record at periods $t = 1, \dots, T$ and $R_t$
the recruits scored at a subset of periods. The lag hypothesis, in the form
a distributed-lag or antecedent-effect model estimates and in the form a
projection matrix asserts, is that expected recruitment is a weighted sum of
past reproduction,

$$
\mathbb{E}[R_t] \;=\; c \sum_{k=0}^{K} w_k \, X_{t - \ell_0 - k},
\qquad w_k \ge 0,\quad \sum_k w_k = 1 ,
$$

with $c$ a scale, $\ell_0$ the offset of the first bin ($\ell_0 = 1$ is the
projection-matrix convention: the first bin is the census before the recruit
census) and $w$ the lag profile. The projection-matrix null is the special
case $w = (1, 0, \dots, 0)$. `convolve_lag()` computes the sum for a given
$w$; `lag_kernels()` builds families of $w$; the scale $c$ never matters for
what follows.

```{r}
X <- lepanthes_census$inflorescences
convolve_lag(X, c(0, 0, 0, 0, 1), t_R = 3:12)          # pure 24 to 30 month delay
convolve_lag(X, flat_kernel(4), t_R = 3:12)             # equal weights
```

Lags that fall before the first census are read as the first census's value
(`missing = "backfill"`). This asserts that reproduction before the record
began was at the level first observed. The alternative, zero-filling,
asserts that no seed fell before the record began, which turns every long
pure delay into a prediction of no recruits at the first censuses; that is an
artefact of the record's start, not a property of the hypothesis, and the
package does not offer it.

## 2. The ceiling: averaging cannot concentrate

Write the expected series as $\mu = W X$ with $W$ the $n \times T$ matrix of
weights. Each row of $W$ is non-negative and sums to one, so each $\mu_t$ is
a weighted average of the $X$ values in its window. The concentration
indices used here, the Gini coefficient

$$
G(x) = \frac{2 \sum_{i=1}^{n} i\, x_{(i)}}{n \sum_i x_i} - \frac{n+1}{n},
$$

with $x_{(i)}$ the values in increasing order, and the coefficient of
variation $\mathrm{CV}(x) = s_x / \bar x$, are both Schur-convex: they fall
whenever a quantity is moved from a fuller period to an emptier one, and
they are invariant to scale. By the Hardy, Littlewood and Pólya theorem, a
vector obtained from another by a doubly stochastic matrix is majorized by
it, and every Schur-convex index of the output is therefore no larger than
that of the input. Over a full record, in which every $X_t$ enters the same
number of windows with the same total weight, $W$ is doubly stochastic up to
the boundary, and the most concentrated output available is the pure delay,
which permutes the series without mixing it. Hence

$$
\text{concentration}\big(\mathbb{E}[R]\big) \;\le\; \text{concentration}(X),
$$

with equality at a pure delay. Reproduction's own concentration is the
ceiling. `lag_ceiling()` reports it as `theorem` and confirms it by searching
the kernel families on exactly the periods at which recruits are scored; the
searched maximum can differ from the bound by a small amount in either
direction because of the boundary (backfilling) and because the scored
periods are a subset of the record.

```{r}
ce <- lag_ceiling(lepanthes_census, reproduction = "inflorescences", K = 4,
                  kernels = lag_kernels(4, n_dirichlet = 2000, bin = 6, unit = "mo"))
ce$theorem; ce$ceiling; ce$best$label
```

The searched cloud shows the whole family sitting below the bound, with
pure delays at its edge:

```{r}
s <- ce$search
plot(s$gini, s$cv, pch = 16, cex = 0.5, col = ifelse(s$family == "pure delay", "#D55E00", "#56B4E9"),
     xlab = "Gini of the expected series", ylab = "CV of the expected series",
     xlim = c(0, 0.7), ylim = c(0, 1.5))
abline(v = ce$theorem["gini"], h = ce$theorem["cv"], lty = 2)
points(ce$obs["gini"], ce$obs["cv"], pch = 17, cex = 1.6)
legend("topleft", bty = "n", pch = c(16, 16, 17), col = c("#56B4E9", "#D55E00", "black"),
       legend = c("searched kernels", "pure delays", "observed recruitment"))
```

The bound is stated for the expectation. The observed series is a count
about that expectation, and counts are more concentrated than their mean.
The exceedance `ce$exceedance` is therefore a description, not a test.

## 3. The second diagnosis: the lag is unidentifiable

If $X$ is nearly constant, $W X$ is nearly constant for every $W$, and the
data carry almost no information about $w$. A distributed-lag model fitted
to such data returns its prior, however long the series runs; the fitted
profile in `lepanthes$lag_profile` is close to flat for that reason.
`rain_expected()` makes the point visually: profiles that could not differ
more give expected series that all but coincide.

## 4. Calibrating the pooled comparison

To ask whether an exceedance of the ceiling by counts means anything, take a
mean that obeys the ceiling (the flat mean, or the best-kernel mean rescaled
to the observed total) and draw negative-binomial counts about it with the
observed variance-to-mean ratio $v$, so that the size parameter is
$\phi = \mu / (v - 1)$. The probability that the simulated series is at
least as concentrated as the observed one is what `ceiling_calibration()`
reports. In the *Lepanthes* record and in every invented record of
`regimes`, that probability is ordinary: pooled over units, small clumped
counts about a flat mean look episodic whatever the mechanism.

## 5. The host-level null, stated as the hypothesis is asserted

The lag hypothesis is a statement about each unit. For unit $p$ with recruit
totals $S_p = \sum_t R_{p,t}$, the null fixes

$$
\mu_{p,t} \;=\; S_p \,\frac{\sum_k w_k X_{p,\,t-\ell_0-k}}{\sum_{t'} \sum_k w_k X_{p,\,t'-\ell_0-k}},
$$

so that the shape of each unit's expected series is its own lagged
reproduction and its scale is its own observed total. Nothing is fitted. A
unit with no lagged reproduction on record but with recruits is given a flat
expectation; a unit that never recruited contributes nothing.
`expected_recruits()` returns this matrix.

Clumping is estimated under that heterogeneous mean. For negative-binomial
counts with means $\mu_{p,t}$ and common size $\phi$,
$\mathbb{E}[(R - \mu)^2] = \mathbb{E}[\mu] + \mathbb{E}[\mu^2]/\phi$, so

$$
\hat\phi \;=\; \frac{\overline{\mu^2}}{\overline{(R-\mu)^2} - \bar\mu}
$$

over the units that recruited (`phi_moment()`). Estimating $\phi$ about a
single pooled mean instead attributes the between-unit spread of the means
to clumping and roughly halves $\hat\phi$; the companion paper reports both,
with the single-mean value as a stress test.

```{r}
h <- host_matrices(lepanthes_hosts, unit = "host", reproduction = "inflorescences")
mu <- expected_recruits(h$R, h$X, w = c(0, 0, 0, 0, 1))
rowSums(mu)[1:5]; rowSums(h$R)[1:5]     # scaled to each host's total
phi_moment(h$R, mu)
```

`host_matrices()` is what the data-frame interface of `expected_recruits()`
and `host_lag_test()` calls; a long tibble with `unit`, `period`,
`reproduction` and `recruits` columns can be passed directly.

`host_lag_test()` then draws, for each profile, `nsim` negative-binomial
count matrices about $\mu$ with size $\hat\phi$, sums them over units to a
census series, and computes four statistics of that series: the number of
silent censuses (no recruit on any unit), the Gini and CV of the census
totals, and the largest total. The reported probability is the fraction of
draws at least as extreme as the observed record. Over profiles, the verdict
is the maximum of each probability, that is, the best case the lag
hypothesis can make for itself. The Gini and CV are the pre-specified
statistics of the paper; the silent-census count is the most interpretable
where a gate closes completely, and the largest census the least sensitive.

## 6. What the test does and does not say

A small verdict says that no delayed consequence of each unit's own
reproduction, of any profile over the horizon searched, produces a record as
concentrated as the observed one: the timing of recruitment is not set by
the timing of seed supply. It does not identify what sets it. An ordinary
verdict says the record is consistent with a delay; it does not identify the
delay, which under continuous reproduction cannot be recovered (Section 3).
Where reproduction is seasonal the ceiling is high and the test is weak, as
the invented records show; aseasonality is where it bites.

## Function map

| Step | Function | Companion-paper phase |
|---|---|---|
| Concentration indices | `rain_gini()`, `rain_cv()`, `concentration()` | Table of series |
| Lag profiles | `lag_kernels()`, `flat_kernel()` | 9a, 9d |
| From a long data frame to series and matrices | `series_from()`, `host_matrices()` | data preparation |
| Expected series | `convolve_lag()` | 9a, 9b |
| Ceiling and search | `lag_ceiling()` | 9a |
| Pooled calibration | `ceiling_calibration()` | 9b |
| Unit-level expectation and clumping | `expected_recruits()`, `phi_moment()` | 9d |
| Host-level test | `host_lag_test()` | 9d |
| Figures | `rain_series()`, `rain_strip()`, `rain_profiles()`, `rain_expected()`, `plot()` methods | Figs 2, 3 and 5 |
